---
title: "Services trade and the Indonesian third unbundling"
author:
  - name: Krisna Gupta
    orcid: 0000-0001-8695-0514
    email: krisna.gupta@cips-indonesia.org
    affiliations:
    - name: Center for Indonesian Policy Studies, Politeknik APP Jakarta
  - name: Dhany S. Ratana
    affiliations: Politeknik APP Jakarta
date: last-modified
date-format: long
bibliography: ref.bib
abstract: "Awokowkoawkoawk."
abstract-title: "abstract"
format:
  html:
    code-fold: true
  pdf:
    documentclass: scrreprt
    papersize: a4
  docx:
    toc: false
---


## Introduction

Trade in services is growing in importance all over the world, Indonesia included. @fig-1 shows Indonesia's export and import in services taken from @seki. Indonesia's services trade is growing steadily from 2000, only to be interrupted by the COVID-19 pandemic.  Export service is dominated by tourism, while import services is dominated by logistics. While the trend is increasing, it is evident that Indonesia's import of services has always been dominating exports.

```{r}
#| warning: false
#| message: false
#| echo: false
## Packages!

library(tidyverse)
library(readxl)
library(scales)
```

```{r}
#| label: fig-1
#| fig-cap: "wkwkw"
read_excel("data/services.xlsx") |>
  pivot_longer(!year,names_to="flow",values_to="value") |>
  ggplot(aes(x=year,y=value,color=flow))+geom_line(linewidth=1.1)+
  scale_y_continuous(labels = scales::comma)+
  scale_color_discrete(labels=c('Import', 'Export'))+     
  labs(x="",y="Million current USD",caption="source: SEKI")+theme_classic()
  
```

Indonesian government often concerned with deficit trade, but trade in services has often neglected in the discussion. Indeed, trade balance in goods are often far outweight the deficit in its services counterpart. However, while Indonesia's trade balance fluctuates along with commodity prices and global demand in general, services trade deficit is consistent. With the increasing role of services in the global trade, the deficit looks to be even more important in Indonesia's current account in the future.

The importance of trade in services goes beyond current account. With the ever decreasing cost of trade, separating a value up to tasks level (i.e., the third unbundling) is on the horizon [@baldwin1;@kimura1]. Feedback mechanism from the third unbundling may benefits domestic manufacturing [@kimura1]. Therefore, services trade may be important in the next stage of globalization.

This chapter have at least two objectives. First, we explores the general trade in services in Indonesia. We use BaTIS data [@batis1;@batis2] to show Indonesia's most important services trade and country partners for both export and import. 

Secondly, we investigate the potency of the feedback mechanism from the third unbundling ala @kimura1. That is, we look at how much imported services are embedded in Indonesia's manufacturing sectors aggregated into ICIO classifications. We do this in two ways. First, we use ARDL [@pesaran] to see whether services imports cointegrate with manufacturing exports and GDP. Secondly, we utilize Inter-Country Input-Output (ICIO) data from OECD [@icio] to look at the importance of services for Indonesian manufacturing. 




## Data and Method

- The OECD Inter-Country Input-Output (ICIO), which has 45 unique industries based on ISIC Revision 4 (OECD, 2023).  The database shows how much sectoral value added, both foreign and domestic, that is used by a certain industry. 
- The OECD Services Trade Restrictiveness Index can also be used to either summarize restriction in trade services, especially Indonesia’s trade partner, also can be use as a distance metric in a gravity-model-fashion.
- The OECD International Trade in Services which is self explanatory.
CEPII for gravity variables.

- Literature studies, visualization and descriptive stats
- Describing the pattern of services trade in Indonesia
- Discussing @krisna which find the inability of IJEPA to foster improvement in trade & services and the lesson for the next trade agreements.
IO table @baumol
- Focusing on the manufacturing ISIC rev. 4 (10-27)
- Using input-output to find multiplier of imported goods in those ISIC. 
- Services gravity model [@kimura2;@krisna]
Using PPML to look at the pattern of Indonesian service trade.

## Discussions

- A deep dive on the Indonesian trade in services.
- Provide an insight on the importance of imported and exported services on the Indonesian GVC.
- Policy lessons and possible insight on what to look for the next trade agreements to improve trade in services.



## References {.unnumbered}