---
title: "Services trade and the Indonesian third unbundling"
author: "Krisna Gupta"
institute: "Politeknik APP Jakarta, Center for Indonesian Policy Studies"
date: last-modified
date-format: long
bibliography: ref.bib
abstract: "Awokowkoawkoawk."
abstract-title: "executive summary"
format:
  html:
    code-fold: true
  pdf:
    documentclass: scrreprt
    papersize: a4
  docx:
    toc: false
---


## Introduction

Indonesia has always been a net-importer on services. Export service is dominated by tourism, while import services is dominated by logistics. Indonesian government often concerned with deficit trade, but trade in services has often neglected in the discussion.

```{r}
## Packages!

library(tidyverse)
library(readxl)
library(scales)
```

```{r}
#| label: fig-1
#| fig-cap: "wkwkw"
read_excel("data/services.xlsx") |>
  pivot_longer(!year,names_to="flow",values_to="value") |>
  ggplot(aes(x=year,y=value,color=flow))+geom_line(linewidth=1.1)+
  scale_y_continuous(labels = scales::comma)+
  scale_color_discrete(labels=c('Export', 'Import'))+     
  labs(x="",y="Million current USD")+theme_classic()
  
```

@fig-1 shows us how fucked up Indonesia is.

- With the ever decreasing cost of trade, separating a value up to tasks level (i.e., the third unbundling) is on the horizon [@baldwin1;@kimura1]

- Feedback mechanism from the third unbundling may benefits domestic manufacturing [@kimura1]
- Services trade may be important in the next stage of globalization.
Easing the barrier to services trade, therefore, will be very important for countries like Indonesia [@kimura1;@baldwin2]
- How important is trade in services these days in the Indonesian economy?

## Objective

- Giving an overview of the Indonesian trade in services.
    - By sectors, by partners.
- Investigating the potency of the feedback mechanism from the third unbundling ala @kimura1.
    - That is, see the multiplier effect of import services especially in the manufacturing sector.
- Provide a policy discussions regarding some specific issues depending on point 2 results.
    - Probably will be a more nuanced, Indonesian-context policy discussion version of @kimura1.

## Data and Method

- The OECD Inter-Country Input-Output (ICIO), which has 45 unique industries based on ISIC Revision 4 (OECD, 2023).  The database shows how much sectoral value added, both foreign and domestic, that is used by a certain industry. 
- The OECD Services Trade Restrictiveness Index can also be used to either summarize restriction in trade services, especially Indonesia’s trade partner, also can be use as a distance metric in a gravity-model-fashion.
- The OECD International Trade in Services which is self explanatory.
CEPII for gravity variables.

- Literature studies, visualization and descriptive stats
- Describing the pattern of services trade in Indonesia
- Discussing @krisna which find the inability of IJEPA to foster improvement in trade & services and the lesson for the next trade agreements.
IO table @baumol
- Focusing on the manufacturing ISIC rev. 4 (10-27)
- Using input-output to find multiplier of imported goods in those ISIC. 
- Services gravity model [@kimura2;@krisna]
Using PPML to look at the pattern of Indonesian service trade.

## Discussions

- A deep dive on the Indonesian trade in services.
- Provide an insight on the importance of imported and exported services on the Indonesian GVC.
- Policy lessons and possible insight on what to look for the next trade agreements to improve trade in services.



## References {.unnumbered}